---
title: "Imperfect detection - assignment"
author: ""
date: "2023-12-03"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
#setwd('C:/Users/johns/OneDrive/Documents/UNIVERSITY/Year 3/Computing/assignment_imperfectdetection_r/assignment') #would have to change working directory if code is run on a different computer.
#knitr::opts_chunk$set(echo = TRUE)
#install.packages('unmarked')
#install.packages('AHMbook')
library(unmarked, AHMbook)
```


# Question 1: Form meaningful biological hypotheses about how Opossum's occupancy and detection probabilities might vary

**Key information:**

 + Opossums are sensitive to temperature
 
 + Opossums are generalist in terms of habitat preference
 
 + Opossum activity patterns are affected by fire

**Biological hypotheses:**

* Hypothesis 1:

   + *Opossum occupancy increases with average yearly temperature*

* Hypothesis 2:

   + *Opossum occupancy does not vary with habitat quality*

* Hypothesis 3:

   + *Opossum detection probability increases with occurrence of fire*
   
--------------------------------------------------------------------------------



# Question 2: Prepare the data and run the models corresponding to your biological hypotheses from Q1

## Prepare the data

```{r}
#Load opossum data

opossum = read.csv("opossumdata.csv")
opossum

```


```{r}
#Separate detection history

y=opossum[,2:9]
y

```

```{r}
#Organise co-variates

siteCovs = data.frame(Temperature=opossum$Temperature, Habitat=opossum$Habitat)
obsCovs = data.frame(opossum[,12:19])
obsCovs = list(obsCovs=obsCovs)
siteCovs
obsCovs

```

```{r}
#Organise data using UMF

umf = unmarkedFrameOccu(y = y, siteCovs = siteCovs, obsCovs = obsCovs)
summary(umf)

```

```{r}
#Standardise the covariates

umf@siteCovs$Temperature = scale(umf@siteCovs$Temperature)

umf@siteCovs$Habitat = scale(umf@siteCovs$Habitat)

umf@obsCovs$obsCovs = scale(umf@obsCovs$obsCovs)

```



## Create the models

**Model 1**

```{r}
#Model 1 - no covariates

model1 = occu(formula = ~1 ~1, data = umf)

model1

#backtransform results from LOGIT scale

backTransform(model1, type="det") #(detection probability, p = 0.530)
backTransform(model1, type="state") #(occupancy probability, p = 0.952)

```


**Model 2**

```{r}
#Model 2 - No detection covariate, occupancy covariate of average yearly temperature ('Temperature')

model2 = occu(formula = ~1 ~Temperature, data = umf)

model2

#backtransform results from LOGIT scale

backTransform(model2, type="det") #(detection probability, p = 0.531)

lincombA1 = linearComb(model2, coefficients = c(1,0), type = "state")
backTransform(lincombA1) #(occupancy probability (low temperature), p = 1)

lincombB1 = linearComb(model2, coefficients = c(1,1), type = "state")
backTransform(lincombB1) #(occupancy probability (high temperature), p = 1)
```


**Model 3**

```{r}
#Model 3 - No detection covatiate, occupancy covariate of habitat quality ('Habitat')

model3 = occu(formula = ~1 ~Habitat, data = umf)

model3

#backtransform results from LOGIT scale

backTransform(model3, type="det") #(detection probability, p = 0.531)

lincombA2 = linearComb(model3, coefficients = c(1,0), type = "state")
backTransform(lincombA2) #(occupancy probability (low habitat quality), p = 1)

lincombB2 = linearComb(model3, coefficients = c(1,1), type = "state")
backTransform(lincombB2) #(occupancy probability (high habitat quality), p = 1)

```


**Model 4**

```{r}
#Model 4 - detection covariate of occurence of fire ('obsCovs'), no occupancy covariate


model4 = occu(formula = ~ obsCovs
                        ~ 1,
              data = umf)
model4

#backtransform results from LOGIT scale

lincombA3 = linearComb(model4, coefficients = c(1,0), type = "det")
backTransform(lincombA3) #(detection probability (no fire), p = 0.532)

lincombB3 = linearComb(model4, coefficients = c(1,1), type = "det")
backTransform(lincombB3) #(detection probability (fire), p = 0.584)

backTransform(model4, type="state") #(occupancy probability, p = 0.952)

```
## Compare models 1, 2, 3 and 4 using AIC

```{r}

fit = fitList('psi(.)p(.)' = model1,
              'psi(Temperature)p(.)' = model2,
              'psi(Habitat)p(.)'=model3,
              'psi(.)p(obsCovs)' = model4)

modSel(fit)

```


**Q2 Conclusion Part 1**

 + The best models are Models 2 and 3:
   + Model 2 - AIC = 206.46 - No detection covariate, occupancy covariate of average yearly temperature ('Temperature')
   + Model 3 - AIC = 207.48 - No detection covatiate, occupancy covariate of habitat quality ('Habitat')
   
 + Cannot determine which of Models 2 and 3 are more suitable, as they are within +/- 2 AIC of each other
 

**Q2 Conclusion Part 2** 
 
 + **Hypotheses:**
 
   + Hypothesis 1 - *Opossum occupancy increases with average yearly temperature*
     + tested by Model 2 - hypothesis **is not** supported as occupancy probability was 1 (did not vary) for both low and high average yearly temperatures.
    
   + Hypothesis 2 - *Opossum occupancy does not vary with habitat quality*
     + tested by Model 3 - hypothesis **is** supported as occupany probability was 1 (did not vary) for low and high habitat quality
     
   + Hypothesis 3 - *Opossum detection probability increases with occurrence of fire*
     + tested by Model 4 - hypothesis **is** supported as detection probability was 0.532 when there was no fire, but it was 0.584 (increased) when there was fire


--------------------------------------------------------------------------------



# Question 3: Interpret the obtained (back transformed) values of the estimates for occupancy and detection probability from the best ranked model(s)

**The best ranked models:**

 + Model 2:

   + Occupancy probability was 1 for low and high average yearly temperatures. This shows that **these data** suggest that temperature (within the 13C to 42.3C range) does not influence Opossum occupation, the probability of occupation is the same (1) for all temperatures

   + Detection probability was 0.531. This gives the probability of detecting an individual in each survey given that that individual does occupy the region being studied.
   

 + Model 3:

   + Occupancy probability was 1 for all habitat quality. This shows that **these data** suggest that habitat quality does not influence Opossum occupation, as the probability of occupation is the same (1) for all habitat qualities

   + Detection probability was 0.531. This gives the probability of detecting an individual in each survey given that that individual does occupy the region being studied.
   
--------------------------------------------------------------------------------

# Question 4: Use the dodo data file. Add the habitat quality covariate. Build a set of single-season models to test the validity of the following hypotheses:
### - *Dodo's occupancy probability is constant while detection probability varies with habitat quality*
### - *Dodo's occupancy probability varies with habitat quality and detection probability varies with occurrence of flood*

$~$
$~$
$~$

**Plan** - Models to build:

Model A: constant occupancy probability (~1), but detection probability has habitat quality covariate

Model B: occupancy probability has habitat quality covariate and detection probability has flood covariate

- Compare Model A and Model B using AIC

$~$
$~$
$~$

**Prepare Dodo data**

```{r}
# Load edited dodo data

dodo_edit = read.csv("Dodo_data_edited.csv")
dodo_edit

```

```{r}
#Separate detection histories

y=dodo_edit[,2:10]
y[y>1]=1
y

```


```{r}
#Organise covariates

siteCovs = data.frame(Habitat=dodo_edit$Habitat, Sun=dodo_edit$Sun)
obsCovs = data.frame(dodo_edit[,13:21])
obsCovs = list(obsCovs=obsCovs)
siteCovs
obsCovs

```

```{r}
#Organise data using UMF

dodo_umf = unmarkedFrameOccu(y=y, siteCovs=siteCovs, obsCovs=obsCovs)
summary(dodo_umf)

```

```{r}
#Standardise covariates

dodo_umf@siteCovs$Habitat = scale(dodo_umf@siteCovs$Habitat)
dodo_umf@siteCovs$Sun = scale(dodo_umf@siteCovs$Sun)
dodo_umf@obsCovs$obsCovs = scale(dodo_umf@obsCovs$obsCovs)



```

$~$
$~$
$~$

**Create models**

**ModelA** - constant occupancy probability (~1), but detection probability has habitat quality covariate

```{r}

modelA = occu(formula = ~ Habitat 
                        ~ 1,
              data = dodo_umf)

modelA

#backtransform results from LOGIT scale

lincombA4 = linearComb(modelA, coefficients = c(1,0), type = "det")
backTransform(lincombA4) #when there is low quality habitat (detection probability, p = 0.401)

lincombB4 = linearComb(modelA, coefficients = c(1,1), type = "det")
backTransform(lincombB4) #when there is high quality habitat (detection probability, p = 0.498)

backTransform(modelA, type="state") #(occupancy probability, p = 0.931)


```


$~$
$~$
$~$

**Model B** - occupancy probability has habitat covariate and detection probability has flood covariate

```{r}

modelB = occu(formula = ~ obsCovs 
                        ~ Habitat,
              data = dodo_umf)

modelB

#backtransform results from LOGIT scale

lincombA5 = linearComb(modelB, coefficients = c(1,0), type = "det")
backTransform(lincombA5) #when there is no flood (detection probability, p =0.423)

lincombB5 = linearComb(modelB, coefficients = c(1,1), type = "det")
backTransform(lincombB5) #when there is flood (detection probability, p = 0.336)

lincombC5 = linearComb(modelB, coefficients = c(1,0), type = "state")
backTransform(lincombC5) #when there low habitat (occupation probability, p =0.942)

lincombD5 = linearComb(modelB, coefficients = c(1,1), type = "state")
backTransform(lincombD5) #when there high habitat (occupation probability, p = 0.983)



```

**Compare Model A and Model B's validity using AIC**

```{r}
fit = fitList('psi(.)p(Habitat)' = modelA,
              'psi(Habitat)p(obsCovs)' = modelB)

modSel(fit)
```
**Question 4 Conclusions:**

 + AIC values:
   + Model A - 108.98
   + Model B - 109.50
 + No significant difference of fit for Models A and B, as AIC scores are within +-2 of each other

Model A: **Model A shows that quality of habitat increases detection probability**

 + Detection Probability (low quality habitat) = 0.401

 + Detection Probability (high quality habitat) = 0.498

 + Occupancy Probability = 0.931
 

Model B: **Model B shows that prescence of flood decreases detection probability and quality of habitat increases occupation probability**

 
 + detection probability (no flood) = 0.423

 + detection probability (with flood) = 0.336

 + occupation probability (low habitat quality) = 0.942

 + occupation probability (high habitat quality) = 0.983
 
--------------------------------------------------------------------------------

\newpage


# Question 5: Build a single-season model to investigate the validity that the occurrence of flood might affect dodo's occupancy probability

**Create model**

Model C: detection probability covariate (~1) and occupation covariate (obsCovs)

```{r}

#modelC = occu(formula = ~ 1 
#                        ~ obsCovs,
#              data = dodo_umf)

#modelC

```

Code chunk causes error.

**Why do you get an error?**

You get an error as occurrence of flood cannot be applied as an occupancy covariate, because the single-season imperfect detection model assumes that occupancy of a region does not vary within a season. So as the flood covariate varies within the season it would only influence occupancy of a site within the season, so this causes an error as it contradicts the model assumption that occupancy of a region cannot change within a season. An occurrence covariate must be constant within a site throughout the season, although it can change between sites. For example, habitat quality of a site does not change within a season, but different sites may have different habitat qualities.



















